Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-03-03 writing/manuscript.Rmd
updated results and discussion following last fix

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig. S5), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs. S4-S6). <!--The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range. -->Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. S9). All species adapted their daily and seasonal movement behavior to changes in temperature (Fig. \@ref(fig:dist)). Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C. However, uncertainties around the estimated effects were higher at extreme temperatures due to data scarcity.

All species' movement frequency and speed decreased with increasing sampling interval (Fig. S10). While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. Consequently, we suggest being cautious when comparing estimated movement behaviors across species even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{st}$ (\\textbf{A}) and day of year at 12:0 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover. Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Estimated RSS values were generally most uncertain at extreme temperatures (Fig. S11).

@@

\noindent The direction and magnitude of change in distance traveled varied greatly between species, but worse climate-change scenarios generally corresponded to greater absolute changes in behavior (Fig. \@ref(fig:t-dist)). Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while under the worst-case scenario changes ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou). Elk and boreal caribou showed relatively little projected changes in yearly distance traveled. Throughout BC (Fig. \@ref(fig:s-dist)), all species but elk and boreal caribou were expected to change their movement rates by 10 to 20% in 2100, relative to 2025, but the predicted changes varied spatially due to heterogeneity in climate change throughout the province. Across all species, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario.

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 0.025 and 0.975 quantiles of the estimated means. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

@@

All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario, but changes were often were only between -1% and -3% under the best-case scenario. Under the worst-case scenario, changes in RSS ranged from -8% to -14% (\@ref:fig(t-hrsfs)). As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range. However, this was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent. Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

@@

\noindent We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates. However, as habitats warm and animals are exposed to increasingly frequent extreme and novel conditions, our ability to confidently predict mammals' responses decreases. Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021]. Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig. \@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].

***continue reading from HERE***

Our ability respond to current and future changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behavior becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behavior due to climate change. An important limitation of Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) is that we do not include measures of uncertainty around the estimated responses, as they cannot be incorporated easily in surface plots, and the purpose of the figures was to illustrate the magnitude and direction of the responses rather than contrasting the surfaces against some hypotheses. Still, we stress that uncertainty is crucial in producing transparent and informed results, particularly when using such estimates for decision-making or predicting future changes, as we do in Figs. \@ref(fig:t-dist) and \@ref(fig:t-hrsfs). We present the spatial estimates of distance traveled and relative habitat selection strength in 2100 without measures of uncertainty for conciseness, but we suggest that future studies and reports include some form of compatibility intervals [@gelman_are_2019], whether these be Bayesian Credible intervals [@mcelreath_statistical_2020] or Frequentist Confidence Intervals, although the latter do not provide a measure of uncertainty in the common sense of the word [@morey_fallacy_2016; @amrhein_scientists_2019]. As alternatives to heat-map surface plots, one may include compatibility intervals using faceted or perspective plots of the surfaces along with the intervals [e.g., Fig. 5 in @akkaya_hocagil_benchmark_2024] or overlapping contour plots on the surfaces [e.g., Fig. 4.14 on page 187 of @wood_generalized_2017]. For a simpler figure, one could also include lines with compatibility intervals for the responses at different temperatures.

@@

\noindent The lack of any common effects of temperature across all species' movement rates indicates that the effects of climate change will vary across species and that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways. Although our models do not account for any physiological or explicit phenological changes (e.g., in the timing of sleep and hibernation), the tensor product terms in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns. For example, at cold temperatures (-20$^\circ$C), elk moved most in the early morning (6:00 PDT) and evening (18:00 PDT), but they moved almost four times more at 6:00 PDT and half as much at 18:00 PDT when temperatures were above 20$^\circ$C. This suggests that while elk maintain a preference for crepuscular activity, they adapt their movement behavior to move more during warmer periods of the day (i.e., evening) on cold days and during cooler periods (i.e., early morning) on warmer days. The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will often have community-wide impacts on movement rates, encounter rates [@martinez-garcia_how_2020], and potentially community structure. More work is necessary on interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022], not only as a direct effect of temperature, but also because of cascading changes in the availability and unpredictability of resources [@mezzini_how_2024].

The greater uncertainty around the estimated movement rates at temperatures outside the typical temperatures for a given day (Figs. S2, S4-S6) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior. In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research. We expand on the importance of quantifying and communicating uncertainty in the last section of this discussion. Our relatively simple models indicate that changes in temperature will alter movement behavior even without accounting for large-scale physiological and behavioral adaptations, which may not be possible at the current rate of climate change [@williams_range_2018; @hetem_responses_2014]. Still, our models to indicate that some animals adapt their behavior phenology in response to temperature, as indicated by wolves', elk's, caribou's, and grizzly bears' increased movement rates during warmer winters and spring days.

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021]. The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in figure \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. The width of the averaged credible intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.

The widespread decrease in RSS in for areas with ranges with the top 5% RSS underscores the importance of moving beyond only using central estimates of change. While changes in the median RSS over time are small to negligible (yet never positive), decreases in the top 5% are more accentuated. This perspective is particularly important in the case of habitat selection modeling, since animals will be selecting most for (and thus spending more time in) habitats with high RSS values, so changes in the upper extremes are more important than changes in the median or lower extreme values. Similar arguments can be made for incorporating variation in temperature (including extreme events) when predicting future changes. Estimating changes in future weather with monthly means alone is insufficient because a change of 1-3 in mean monthly temperature may result in negligible predicted changes while ignoring changes in the frequency and intensity of extreme temperatures [@bunnell_global_2011; @zhang_increased_2023; @yao_emergence_2022].

@@

Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024]. The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@need_ref], and food webs [@need_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].

>>>>

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Figs. S5 and S13), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs. S6-S8). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. S9). All species adapted their daily and seasonal movement behavior to changes in temperature (Fig. \@ref(fig:dist)). Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.

All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. S10). While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980; also see @nathan_big-data_2022]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. Consequently, we suggest being cautious when comparing estimated movement behaviors across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

The number of available points greatly outnumbered the number of used points (Fig. S14), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig. S14). Species' relative selection strength (RSS) for resources depended on temperature, and RSS was generally strongest for elevation and weakest for forest cover. Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, and pumas increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Estimated RSS values were generally most uncertain at extreme temperatures (Fig. S13).

@@

\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behavior (Figs. \@ref(fig:t-dist), S11, and S12). Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (pumas), although the models did not explicitly account for changes in hibernation phenology. Throughout BC (Fig. \@ref(fig:s-dist)), all species showed little to no absolute change under the best-case scenario (approximately 0-4% relative to 2025, on average), and approximately 2-25% average absolute change under the worst-case scenario, but the predicted changes varied spatially due to heterogeneity in climate change. Again, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario for all species.

```{r, eval=FALSE, echo=FALSE}
# local predictions
readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Best', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

# BC predictions
readRDS('../data/cc-hgam-bc-projections-albers.rds') %>%
  group_by(lab) %>%
  filter(grepl('Best', scenario)) %>%
  summarise(mean = mean(d)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)

readRDS('../data/cc-hgam-projections.rds') %>%
  group_by(species) %>%
  filter(grepl('Worst', scenario), year > 2095) %>%
  mutate(d_median = d_median / d_ref) %>%
  summarise(mean = mean(d_median)) %>%
  mutate(mean = abs(mean - 1) * 100) %>%
  arrange(mean)
```

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 90% prediction interval within the range. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

@@

All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario. Changes were often were only between -1% and -3% under the best-case scenario but they ranged from -8% to -14% under the worst-case scenario (\@ref:fig(t-hrsfs)). As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range. The increase was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent. Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

@@

\noindent We have demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers allowed us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting at mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021]. Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig. \@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].

Our success in respond to current and future changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behavior becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behavior due to climate change. An important limitation of Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) is that we do not include measures of uncertainty around the estimated responses, as they cannot be incorporated easily in surface plots, and the purpose of the figures was to illustrate the magnitude and direction of the responses rather than contrasting the surfaces against some hypotheses. Still, we stress that uncertainty is crucial in producing transparent and informed results, particularly when using such estimates for decision-making or predicting future changes, as we do in Figs. \@ref(fig:t-dist) and \@ref(fig:t-hrsfs). We present the spatial estimates of distance traveled and relative habitat selection strength in 2100 without measures of uncertainty for conciseness, but we suggest that future studies and reports include some form of compatibility intervals [@gelman_are_2019], whether these be Bayesian Credible intervals [@mcelreath_statistical_2020] or Frequentist Confidence Intervals, although the latter do not provide a measure of uncertainty in the common sense of the word [@morey_fallacy_2016; @amrhein_scientists_2019]. As alternatives to heat-map surface plots, one may include compatibility intervals using faceted or perspective plots of the surfaces along with the intervals [e.g., Fig. 5 in @akkaya_hocagil_benchmark_2024] or overlapping contour plots on the surfaces [e.g., Fig. 4.14 on page 187 of @wood_generalized_2017]. For a simpler figure, one could also include lines with compatibility intervals for the responses at different temperatures.

@@

\noindent The lack of common effects of temperature across all species' movement rates indicates that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways. Although our models do not account for any physiological or explicit phenological changes (e.g., in the timing of sleep and hibernation), the tensor product terms in Fig. \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in pumas and grizzly bears). For example, when temperatures were above 0$^\circ$C, pumas moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day. Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer. The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when summer temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020]. More work is necessary on interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024]. As indicated by grizzly bears' tendency to travel greater distances earlier in the year with increasing temperatures, the increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022]. This change will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2024].

The greater uncertainty around the estimated movement rates at temperatures outside the typical temperatures for a given day (Figs. S4, S6-S8) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior. In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research. We expand on the importance of quantifying and communicating uncertainty in the last section of this discussion. Our relatively simple models indicate that changes in temperature will alter movement behavior even without accounting for large-scale physiological and behavioral adaptations, which may not be possible at the current rate of climate change [@williams_range_2018; @hetem_responses_2014]. Still, our models to indicate that some animals adapt their behavior phenology in response to temperature, as indicated by wolves', elk's, caribou's, and grizzly bears' increased movement rates during warmer winters and spring days.

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021]. The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in figure \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. The width of the prediction intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.

@@

Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024]. The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
